Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.

### 2024-04-30 5_artigos/communication-in-statistics/submission/SOURCE/paper htmcglm/main.tex
inclui artigo smj

% sage_latex_guidelines.tex V1.20, 14 January 2017

\documentclass[Review,sagev,times, doublespace]{sagej}

\usepackage{moreverb,url}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2016}

\begin{document}

\runninghead{Freitas et al}

\title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.}

\author{Lineu Alberto Cavazani de Freitas\affilnum{1}, Ligia de Oliveira Carlos\affilnum{2}, Antônio Carlos Ligocki Campos\affilnum{2} and Wagner Hugo Bonat\affilnum{3}}

\affiliation{
    \affilnum{1}Department of Informatics, Paraná Federal University, Paraná, Brazil\\
    \affilnum{2}Department of Surgical Clinic, Paraná Federal University, Paraná, Brazil\\
    \affilnum{3}Department of Statistics, Paraná Federal University, Paraná, Brazil\\    
}

\corrauth{Lineu Alberto Cavazani de Freitas, Department of Informatics, Paraná Federal University, R. Evaristo F. Ferreira da Costa, 383-391, Jardim das Américas, Curitiba, Paraná, 82590-300, Brazil.}

\email{lineuacf@gmail.com}

@@

Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
    The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

\end{abstract}

\keywords{McGLM, Hypothesis tests, Wald test, ANOVA, MANOVA, Multiple comparisons}

\maketitle

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE1.jpeg}
\caption{Exploratory graphic analysis: (a) YFAS histogram, (b) YFAS boxplots as a function of group, (c) YFAS boxplots as a function of moment, (d) BES histogram, (b) BES boxplots as a function of group, (c) BES boxplots as a function of moment. The asterisk in the boxplots indicates the mean.\label{fig1}}
\end{figure}

@@

For each point, we evaluated the percentage of rejection of the null hypothesis. The idea is to verify what happens when we move away from the null hypotheses. It is expected that at the first point there will be a low rejection rate, since the null hypothesis corresponds to the real values of the parameters. For the other points, it is expected that the rate of rejection will increase gradually, as the hypotheses move away and further away from the originally simulated values. The hypotheses tested in each scenario are available in the appendix.

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE2.jpeg}
\caption{Simulation study results for the regression parameters.\label{fig2}}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE3.jpeg}
\caption{Simulation study results for the dispersion parameters.\label{fig3}}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE4.jpeg}
\caption{Histogram of Pearson residuals by response.\label{fig4}}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE5.jpeg}
\caption{Pearson residuals versus predicted plot with smooth trend line for each response.\label{fig5}}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{c|cccccc}
\hline
\multirow{2}{*}{Parameter} & \multicolumn{3}{c}{YFAS}                                                                                             & \multicolumn{3}{c}{BES}                                                                         \\ \cline{2-7} 
                           & Estimate & \begin{tabular}[c]{@{}c@{}}Confidence \\ interval\end{tabular} & \multicolumn{1}{c|}{p-value}        & Estimate & \begin{tabular}[c]{@{}c@{}}Confidence \\ interval\end{tabular} & p-value        \\ \hline
$\beta_0$                  & -0.54      & (-0.87;-0.22)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.13      & (-1.44;-0.83)                                                     & \textless 0.01 \\
$\beta_1$                  & -1.55      & (-2.17;-0.94)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.16      & (-1.62;-0.69)                                                     & \textless 0.01 \\
$\beta_2$                  & -1.13      & (-1.75;-0.51)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.05      & (-1.58;-0.52)                                                     & \textless 0.01 \\
$\beta_3$                  & 0.49       & (0.05;0.93)                                                       & \multicolumn{1}{c|}{0.0284} & 0.37       & (-0.03;0.77)                                                      & 0.0733           \\
$\beta_4$                  & -0.73      & (-1.60;0.14)                                                      & \multicolumn{1}{c|}{0.0}            & -0.33      & (-0.96;0.30)                                                      & 0.3081           \\
$\beta_5$                  & -0.98      & (-1.93;-0.03)                                                     & \multicolumn{1}{c|}{0.0429} & -0.80      & (-1.58;-0.02)                                                     & 0.0449 \\
$\tau_0$                   & 0.18       & (0.01;0.35)                                                       & \multicolumn{1}{c|}{0.0411} & 0.17       & (0.00;0.34)                                                       & 0.0458           \\
$\tau1$                    & 0.01       & (-0.02;0.04)                                                      & \multicolumn{1}{c|}{0.5718}           & 0.04       & (-0.01;0.10)                                                      & 0.1357           \\
$p$                        & 0.91       & (0.47;1.34)                                                       & \multicolumn{1}{c|}{\textless 0.05} & 1.23       & (0.77;1.68)                                                       & \textless 0.05 \\ \hline
\end{tabular}
\caption{Parameter estimates, 95\% confidence intervals and p-values of the model.}
\label{tab3}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=31.5pc,height=18pc]{FIGURE6.jpeg}
\caption{Graph of model predicted for each time and group combination.\label{fig6}}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{cccc}
\hline
Variable      & Degrees of freedom & Chi-square   & p-value        \\ \hline
Intercept     & 2                  & 53.1581      & \textless 0.01 \\
Moment        & 8                  & 139.0161     & \textless 0.01 \\
Group         & 6                  & 8.4928       & 0.2042         \\
Moment*Group  & 4                  & 6.9923       & 0.1363         \\ \hline
\end{tabular}
\caption{Type II multivariate analysis of variance.}
\label{tab4}
\end{table}

@@

\begin{table}[h]
\centering
\begin{tabular}{c|c|cc|cc}
\hline
              &                    & \multicolumn{2}{c|}{YFAS}     & \multicolumn{2}{c}{BES}       \\ \hline
Variable      & Degrees of freedom & Chi-square   & p-value        & Chi-square   & p-value        \\ \hline
Intercept     & 1                  & 10.6128      & \textless 0.01 & 53.1473      & \textless 0.01 \\
Moment        & 4                  & 102.9875     & \textless 0.01 & 99.5681      & \textless 0.01 \\
Group         & 3                  & 6.6837       & 0.0827         & 5.3083       & 0.1506         \\
Moment*Group  & 2                  & 5.5984       & 0.0609         & 4.2477       & 0.1196         \\ \hline
\end{tabular}
\caption{Type II univariate analysis of variance.}
\label{tab5}
\end{table}

@@

\begin{table}[h]
\centering
\begin{tabular}{cccc}
\hline
Contrast  & Degrees of freedom & Chi-square   & p-value        \\ \hline
T0-T1     & 2                  & 97.9874      & \textless 0.01 \\
T0-T2     & 2                  & 67.2462      & \textless 0.01 \\
T1-T2     & 2                  & 2.4730       & 0.8712         \\ \hline
\end{tabular}
\caption{Two-by-two comparisons between moments for both responses.}
\label{tab6}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{cccc}
\hline
Contrast                 & Degrees of freedom & Chi-square   & p-value \\ \hline
T0:Placebo-T0:Probiotic  & 2                  & 5.5819       & 0.9204  \\
T1:Placebo-T1:Probiotic  & 2                  & 0.6096       & 1       \\
T2:Placebo-T2:Probiotic  & 2                  & 1.7645       & 1       \\ \hline
\end{tabular}
\caption{Two-by-two comparisons between groups for each time point for both responses.}
\label{tab7}
\end{table}

@@

\begin{table}[h]
\centering
\begin{tabular}{cccc}
\hline
Variable               & Degrees of freedom & Chi-square & p-value        \\ \hline
$\tau_0$ & 2                  & 7.1936       & 0.0274  \\
$\tau_1$ & 2                  & 2.3201       & 0.3135         \\ \hline
\end{tabular}
\caption{Type III multivariate analysis of variance for dispersion parameters.}
\label{tab8}
\end{table}

@@

\begin{acks}
The authors thank the reviewers for their constructive and helpful comments, which greatly improved the article.
\end{acks}

%\begin{biog}
%To typeset an
%  "Author biography" section.
%\end{biog}

%\begin{biogs}
%To typeset an
%  "Author Biographies" section.
%\end{biogs}

\begin{dci}
The Authors declares that there is no conflict of interest.
\end{dci}

\begin{funding}
This study was
financed in part by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.
\end{funding}

\begin{sm}
R code and data set are available at https://github.com/lineu96/paper-htmcglm
\end{sm}

\bibliographystyle{SageV}
\bibliography{draft-sm-lineu.bib}

\appendix

>>>>

\documentclass[a4paper]{article}

\usepackage[pages=all, color=black, position={current page.south}, placement=bottom, scale=1, opacity=1, vshift=5mm]{background}
%\SetBgContents{
%	\tt This work is shared under a \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC BY-SA 4.0 license} unless otherwise noted
%}      % copyright

\usepackage[margin=1in]{geometry} % full-width

% AMS Packages
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

% Unicode
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\hypersetup{
	unicode,
%	colorlinks,
%	breaklinks,
%	urlcolor=cyan, 
%	linkcolor=blue, 
	%pdfauthor={Author One, Author Two, Author Three},
	%pdftitle={A simple article template},
	%pdfsubject={A simple article template},
	%pdfkeywords={article, template, simple},
	%pdfproducer={LaTeX},
	%pdfcreator={pdflatex}
}

% Vietnamese
%\usepackage{vntex}

% Natbib
\usepackage[sort&compress,numbers,square]{natbib}
\bibliographystyle{mplainnat}

% Theorem, Lemma, etc
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}{Claim}[theorem]
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{criterion}[theorem]{Criterion}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{principle}[theorem]{Principle}

\usepackage{graphicx, color}
\graphicspath{{fig/}}

%\usepackage[linesnumbered,ruled,vlined,commentsnumbered]{algorithm2e} % use algorithm2e for typesetting algorithms
\usepackage{algorithm, algpseudocode} % use algorithm and algorithmicx for typesetting algorithms
\usepackage{mathrsfs} % for \mathscr command

\usepackage{multirow}
\usepackage{float}

% Author info
\title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.}
\author{Lineu Alberto Cavazani de Freitas$^1$\thanks{Author One was partially supported by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.} \and Ligia de Oliveira Carlos$^2$ \and Antônio Carlos Ligocki Campos$^2$ \and Wagner Hugo Bonat$^3$}

\date{
	$^1$Department of Informatics, Paraná Federal University, Paraná, Brazil\\%
	$^2$Department of Surgical Clinic, Paraná Federal University, Paraná, Brazil\\%
        $^3$Department of Statistics, Paraná Federal University, Paraná, Brazil\\%
%	\today
}

\begin{document}
	\maketitle

@@

\noindent Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
    The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

\vspace{0.5cm}
		\noindent\textbf{Keywords:} McGLM, Hypothesis tests, Wald test, ANOVA, MANOVA, Multiple comparisons
	\end{abstract}

	%\tableofcontents

@@

\begin{figure}[H]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE1.jpeg}
\caption{Exploratory graphic analysis: (a) YFAS histogram, (b) YFAS boxplots as a function of group, (c) YFAS boxplots as a function of moment, (d) BES histogram, (b) BES boxplots as a function of group, (c) BES boxplots as a function of moment. The asterisk in the boxplots indicates the mean.\label{fig1}}
\end{figure}

@@

For each point, we evaluated the percentage of rejection of the null hypothesis. The idea is to verify what happens when we move away from the null hypotheses. It is expected that at the first point there will be a low rejection rate, since the null hypothesis corresponds to the real values of the parameters. For the other points, it is expected that the rate of rejection will increase gradually, as the hypotheses move away and further away from the originally simulated values. The hypotheses tested in each scenario are available in the appendix in \autoref{tab:hipoteses_beta_normal}, \autoref{tab:hipoteses_beta_poisson}, \autoref{tab:hipoteses_beta_bernoulli} and \autoref{tab:hipoteses_taus}.

@@

\begin{figure}[H]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE2.jpeg}
\caption{Simulation study results for the regression parameters.\label{fig2}}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE3.jpeg}
\caption{Simulation study results for the dispersion parameters.\label{fig3}}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE4.jpeg}
\caption{Histogram of Pearson residuals by response.\label{fig4}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE5.jpeg}
\caption{Pearson residuals versus predicted plot with smooth trend line for each response.\label{fig5}}
\end{figure}

@@

\begin{table}[H]
\centering
\begin{tabular}{c|cccccc}
\hline
\multirow{2}{*}{Parameter} & \multicolumn{3}{c}{YFAS}                                                                                             & \multicolumn{3}{c}{BES}                                                                         \\ \cline{2-7} 
                           & Estimate & \begin{tabular}[c]{@{}c@{}}Confidence \\ interval\end{tabular} & \multicolumn{1}{c|}{p-value}        & Estimate & \begin{tabular}[c]{@{}c@{}}Confidence \\ interval\end{tabular} & p-value        \\ \hline
$\beta_0$                  & -0.54      & (-0.87;-0.22)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.13      & (-1.44;-0.83)                                                     & \textless 0.01 \\
$\beta_1$                  & -1.55      & (-2.17;-0.94)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.16      & (-1.62;-0.69)                                                     & \textless 0.01 \\
$\beta_2$                  & -1.13      & (-1.75;-0.51)                                                     & \multicolumn{1}{c|}{\textless 0.01} & -1.05      & (-1.58;-0.52)                                                     & \textless 0.01 \\
$\beta_3$                  & 0.49       & (0.05;0.93)                                                       & \multicolumn{1}{c|}{0.0284} & 0.37       & (-0.03;0.77)                                                      & 0.0733           \\
$\beta_4$                  & -0.73      & (-1.60;0.14)                                                      & \multicolumn{1}{c|}{0.0}            & -0.33      & (-0.96;0.30)                                                      & 0.3081           \\
$\beta_5$                  & -0.98      & (-1.93;-0.03)                                                     & \multicolumn{1}{c|}{0.0429} & -0.80      & (-1.58;-0.02)                                                     & 0.0449 \\
$\tau_0$                   & 0.18       & (0.01;0.35)                                                       & \multicolumn{1}{c|}{0.0411} & 0.17       & (0.00;0.34)                                                       & 0.0458           \\
$\tau1$                    & 0.01       & (-0.02;0.04)                                                      & \multicolumn{1}{c|}{0.5718}           & 0.04       & (-0.01;0.10)                                                      & 0.1357           \\
$p$                        & 0.91       & (0.47;1.34)                                                       & \multicolumn{1}{c|}{\textless 0.05} & 1.23       & (0.77;1.68)                                                       & \textless 0.05 \\ \hline
\end{tabular}
\caption{Parameter estimates, 95\% confidence intervals and p-values of the model.}
\label{tab3}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=31.5pc,height=18pc]{FIGURE6.jpeg}
\caption{Graph of model predicted for each time and group combination.\label{fig6}}
\end{figure}

@@

\begin{table}[H]
\centering
\begin{tabular}{cccc}
\hline
Variable      & Degrees of freedom & Chi-square   & p-value        \\ \hline
Intercept     & 2                  & 53.1581      & \textless 0.01 \\
Moment        & 8                  & 139.0161     & \textless 0.01 \\
Group         & 6                  & 8.4928       & 0.2042         \\
Moment*Group  & 4                  & 6.9923       & 0.1363         \\ \hline
\end{tabular}
\caption{Type II multivariate analysis of variance.}
\label{tab4}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{c|c|cc|cc}
\hline
              &                    & \multicolumn{2}{c|}{YFAS}     & \multicolumn{2}{c}{BES}       \\ \hline
Variable      & Degrees of freedom & Chi-square   & p-value        & Chi-square   & p-value        \\ \hline
Intercept     & 1                  & 10.6128      & \textless 0.01 & 53.1473      & \textless 0.01 \\
Moment        & 4                  & 102.9875     & \textless 0.01 & 99.5681      & \textless 0.01 \\
Group         & 3                  & 6.6837       & 0.0827         & 5.3083       & 0.1506         \\
Moment*Group  & 2                  & 5.5984       & 0.0609         & 4.2477       & 0.1196         \\ \hline
\end{tabular}
\caption{Type II univariate analysis of variance.}
\label{tab5}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{cccc}
\hline
Contrast  & Degrees of freedom & Chi-square   & p-value        \\ \hline
T0-T1     & 2                  & 97.9874      & \textless 0.01 \\
T0-T2     & 2                  & 67.2462      & \textless 0.01 \\
T1-T2     & 2                  & 2.4730       & 0.8712         \\ \hline
\end{tabular}
\caption{Two-by-two comparisons between moments for both responses.}
\label{tab6}
\end{table}

\begin{table}[H]
\centering
\begin{tabular}{cccc}
\hline
Contrast                 & Degrees of freedom & Chi-square   & p-value \\ \hline
T0:Placebo-T0:Probiotic  & 2                  & 5.5819       & 0.9204  \\
T1:Placebo-T1:Probiotic  & 2                  & 0.6096       & 1       \\
T2:Placebo-T2:Probiotic  & 2                  & 1.7645       & 1       \\ \hline
\end{tabular}
\caption{Two-by-two comparisons between groups for each time point for both responses.}
\label{tab7}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{cccc}
\hline
Variable               & Degrees of freedom & Chi-square & p-value        \\ \hline
$\tau_0$ & 2                  & 7.1936       & 0.0274  \\
$\tau_1$ & 2                  & 2.3201       & 0.3135         \\ \hline
\end{tabular}
\caption{Type III multivariate analysis of variance for dispersion parameters.}
\label{tab8}
\end{table}

@@

\paragraph{Acknowledgements:} The authors thank the reviewers for their constructive and helpful comments, which greatly improved the article.

\paragraph{Conflict of interest:} The Authors declares that there is no conflict of interest.

\paragraph{Funding:} This study was financed in part by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.

\paragraph{Data:} R code and data set are available at https://github.com/lineu96/paper-htmcglm

\newpage
\bibliography{refs}

\newpage
\appendix
